\sys: Light-touch Data-Parallel Shell Processing

### 2020-05-22 pldi/draft.tex
Pass over 5

To identify opportunities for parallelism, the translation
pass extracts the parallelizability class of each command together
with its inputs and outputs.  To achieve this for each command,
it searches all available annotation~\sx{ext} and resorts to
conservative defaults if no record is found. If the
command is \sta or parallelizable \pur, the translation pass initiates
a parallelizable region that is propagated up the tree.

@@

An important consideration is that due to the highly dynamic nature of
the shell, there is information (such as the values of environment
variables) that are not known to \sys at the time of translation.
%% to the the safety properties that are mentioned above cannot be
%% soundly checked statically.
For safety purposes, \sys takes a conservative approach where it does
not parallelize nodes without complete information.  For example, the
translation pass cannot infer that an environment variable that is
given as an argument to a command is not a flag that changes its
parallelizability class,
%% two un-expand strings do not refer to the same file identifier,
and as a result does not parallelize such sub-expressions.

@@

After translating the parallelizable regions of the input script to
dataflow graphs and optimizing these graphs using the transformations
described in \Cref{ir:transformations}, \sys implements each graph by
translating it back to a shell script. Nodes of the graph are
instantiated with their respective commands, and edges are
instantiated as named pipes. In this section, we describe several
technical challenges that \sys addresses to efficiently perform this
transformation, as well as the implementations of the custom \sys
nodes that are added by the auxiliary transformations.
%% it instantiates the \todo{custom} split and relay nodes as highly
%% optimized commands to improve In this section we describe how \sys
%% achieves that, and several technical challenges that it addresses.

@@

\heading{Overcoming Laziness}
%
%
The shell is uniquely lazy, in the sense that most commands and pipes
consume their inputs only when they are ready to process more. This
allows the underlying system to concurrently execute more commands
than the available processing threads,
%% support concurrently executing many more commands than the
%% processing threads of the underlying system,
%% by scheduling them concurrently, allowing them to
%% have progress one at a time.
by interleaving their executions. This design decision however,
severely restricts parallelism since nodes are often blocked when
their consumers are not requesting any input. Consider as an example
the following script:

@@

\noindent
The \ttt{cat} command will only try to consume input from \ttt{t2}
when it is done reading from \ttt{t1}. This means that the second
\ttt{grep} will block without processing any input until the first
\ttt{grep} is done processing. Instead of addressing this issue for
each command separately, \sys inserts and instantiates eager
\ttt{relay} nodes in critical points of the pipeline as seen in
\Cref{fig:eager}. These nodes eagerly consume their input, forcing
their upstream nodes to produce output when possible,
%% force their producers to
%% produce output and not block,
while also preserving the existing task-based parallelism---that would
not be preserved say with the \todo{third} solution in
\Cref{fig:eager}.

@@

\begin{figure}[t]
\centering
% https://docs.google.com/drawings/d/13Zx3a9bWi1RvpOgNCSiXirp9UOte-XuaJQBi_lmmPMc/edit
\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_eager.pdf}}
\caption{
  \textbf{Eager primitive.}
  Describe three problems:
    (i) FIFOs alone cannot solve it as they’re blocking;
    (ii) intermediary files alone lose data, as the consumer reaches EOF before the producer; and
    (iii) writing to files + wait does not allow trask-based parallelism.
  Eager is an elegant solution that solves the problem while remaining within
  the model. Describe. \TODO{Combine this figure with the above code maybe.}
}
\vspace{-15pt}
\label{fig:eager}
\end{figure}

\heading{Split node implementation}
%
As described in \Cref{ir:transformations}, \sys's optimizer inserts
\ttt{split} nodes to expose parallelism in cases where a
parallelizable node only has one input. In order for the split to be
effective in exposing parallelism, it needs to split its input equally
in its outputs, but the size of its input can not be known
a-priori. To satisfy this inherent tension, \sys supports two split
implementations, a general one that can be used if the script contains
commands that could add or remove lines, and an optimized one that can
be used if the size of the split input is known before the script
executes.
%% The \sys \todo{backend} support two split node implementations, to
%% satisfy the inherent tension between the split efficiency and
%% task-based parallelism.
In the general case, since the number of lines of a specific edge is
not known, \sys provides an automatic split that first consumes its
complete input, counts its lines, and then splits it equally to the
desired number of outputs. If however, the user knows the number of
lines at the split insertion edge, they can configure \sys to split
without having to read its whole input, therefore exploiting available
task-based parallelism. Finally, \sys also inserts eager relay nodes
after all but the last output of split in order to address the
laziness issue that was described before.

>>>>

To identify opportunities for parallelization, the translation pass extracts each command's  parallelizability class together with its inputs and outputs.
To achieve this for each command, it searches all its available annotations~\sx{ext} and resorts to conservative defaults if none is found.
If the command is in \sta or \pur, the translation pass initiates a parallelizable region that is propagated up the tree.

@@

Due to the highly dynamic nature of the shell, some information is not known to \sys at the time of translation.
Examples of such information includes the values of environment variables, strings that have not been expanded, and sub-shell constructs.
%% to the the safety properties that are mentioned above cannot be
%% soundly checked statically.
For safety purposes, \sys takes a conservative approach avoiding parallelization of such nodes with incomplete information.
It will not attempt to parallelize sub-expressions in which the translation pass cannot infer that, for example, an environment variable passed as an argument to a command does not change its parallelizability class.

@@

After translating the parallelizable regions of the input script to DFGs and applying optimization passes using the transformations described earlier~\sx{ir:transformations},
  \sys translated all DFGs back to a shell script.
Nodes of the graph are instantiated with the commands and flags they represent, and edges are instantiated as named pipes.
This section describes several technical challenges that \sys addresses to efficiently perform this transformation, as well as the implementations of the custom \sys nodes that are added by the auxiliary transformations.
%% it instantiates the \todo{custom} split and relay nodes as highly
%% optimized commands to improve In this section we describe how \sys
%% achieves that, and several technical challenges that it addresses.

@@

\heading{Overcoming Laziness}
The shell's evaluation strategy is unusually lazy, in that most commands and shell constructs consume their inputs only when they are ready to process more.
This allows \unix to multiplex the executing commands efficiently by interleaving their executions. 
%% support concurrently executing many more commands than the
%% processing threads of the underlying system,
%% by scheduling them concurrently, allowing them to
%% have progress one at a time.
However, such laziness severely inhibits opportunities for parallelism as commands are often blocked when their consumers are not requesting any input.
Consider the following script:

@@

\noindent
The \ttt{cat} command will only try to consume input from \ttt{t2} when it completes reading from \ttt{t1}.
As a result, the second \ttt{grep} will remain blocked until the first \ttt{grep} completes. % processing.

% replacing FIFOs with files is
The closest solution \unix provides would be to replace FIFOs with files, as files are a central abstraction and can simulate pipes of arbitrary buffering.
Aside from severe performance implications, naive replacement can lead to subtle race conditions as the consumer might reach EOF before the producer.
Alternatively, a consumer could the producer to complete, but that would insert an artificial barrier inhibiting task-based parallelism and wasting disk resources.
% \unix does not provide an easy way out.

To address this challenge,
% Instead of addressing this issue for each command separately, % nv: how could this be possible?
  \sys inserts and instantiates \ttt{relay} nodes at these points (Fig.~\ref{fig:eager}). % nv: not sure what critical meant
These nodes feature tight multi-threaded loops that consume input eagerly while attempting to push, forcing upstream nodes to produce output when possible while also preserving task-based parallelism.

@@

\begin{figure}[t]
\centering
% https://docs.google.com/drawings/d/13Zx3a9bWi1RvpOgNCSiXirp9UOte-XuaJQBi_lmmPMc/edit
\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_eager.pdf}}
\caption{
  \textbf{Eager primitive.}
  Addressing intermediary laziness is challenging:
    (i) FIFOs are blocking;
    (ii) files alone introduce race conditions between producer/consumer;
    (iii) files \ttt{wait} inhibit task-based parallelism.
Eager \ttt{relay} nodes address the challenge while remaining within the \sys model.
% Describe. \TODO{Combine this figure with the above code maybe.}
}
\vspace{-15pt}
\label{fig:eager}
\end{figure}

\heading{Splitting Challenges}
\sys's optimizer inserts \ttt{split} nodes to expose parallelism when parallelizable nodes only have one input~\sx{ir:transformations}.
For \ttt{split} to be effective, it needs to disperse its input uniformly across its outputs.
This requires the input size to be known beforehand, which is not always the case.

To achieve this, % satisfy the tension between efficiency and \todo{knowledge},  % blanking on the right word here
  \sys supports two \ttt{split} implementations:
  (i) a general one that can be used with commands that could add or remove lines, and
  (ii) an optimized one that can be used if the input size before its execution.
%% The \sys \todo{backend} support two split node implementations, to
%% satisfy the inherent tension between the split efficiency and
%% task-based parallelism.
For the former, \ttt{split} first consumes its complete input, counts its lines, and then splits it equally to the desired number of outputs.
For the latter, it can be configured to avoid reading all its input, thus exploiting available task-based parallelism.

\sys also inserts eager \ttt{relay} nodes after all \ttt{split} outputs (except the last one) to address laziness concerns.
